---
title: "Lab2 - Inference about mean vectors, MANOVA"
author: "Andreas C Charitos [andch552]"
date: "1/22/2021"
header-includes:
 - \usepackage{eso-pic,graphicx,transparent}
output: pdf_document
---

\tableofcontents

\newpage



```{r,,message=FALSE,echo=FALSE}
# Import libraries --------------------------------------------------------
# import libraries --------------------------------------------------------
library(car)
library(latticeExtra)
library(heplots)
library(GGally)
library(knitr)
library(CARS)
```


# Problem 1 


## Data Overview

Table of the first 3 lines of the bird data
-------------------------------------------


```{r, echo=FALSE}
# a) ----------------------------------------------------------------------

birds=read.table("T5-12.DAT")
colnames(birds)=c("Tail Length","Wing Length")
kable(head(birds))

```


## a) 

FInd and sketch the 95% confidence ellipse for the polulation means $\mu_1$ and $\mu_2$. Suppose it is known that $\mu_1=190~mm$ and $\mu_2=275~mm$ for male hook-biled kites. Are these plausible values for the mean tail length and mean wing length for the female birds? Explain.

Confidence Ellipse
------------------

```{r,echo=F}

x1=birds$`Tail Length` ; x2=birds$`Wing Length`
n=dim(birds)[1] ; p=dim(birds)[2]

a=0.05
mu0=c(190,275)
xbar=colMeans(birds)
S=cov(birds)

dist=xbar-mu0
# crit_val=sqrt(p*(n-1)/(n*(n-p)))*qf(1-a,p,n-p)
crit_val=sqrt(p*(n-1)/(n*(n-p))*qf(1-a,p,n-p))


angles=seq(0,2*pi,length.out=200)
# eigen values and eigen vectors of covariance matrix
eigVal <-eigen(S)$values
eigVec <- eigen(S)$vectors
eigScl <- eigVec%*%diag(sqrt(eigVal))
xMat <- rbind(xbar[1] + eigScl[1,]**crit_val, xbar[1]- eigScl[1,]*crit_val)
yMat <- rbind(xbar[2] + eigScl[2,]**crit_val, xbar[2]- eigScl[2,]*crit_val)
ellBase <- cbind(sqrt(eigVal[1])*crit_val*cos(angles), sqrt(eigVal[2])* crit_val*sin(angles))
ellRot <- eigVec%*%t(ellBase)

```


```{r, echo=F}
par(bg='whitesmoke')
plot(birds$`Tail Length`,
     birds$`Wing Length` ,
     xlim = c(170,220),ylim=c(240,320),
     xlab='Tail length',
     ylab='Wing length',
     main='Ellipse plot for 95% confidence',
     panel.first = grid(20,20),pch=20)
lines( (ellRot+xbar)[1,],(ellRot+xbar)[2,],
       asp=1,type="l",lwd=2,col="orange") 
points(xbar[1],xbar[2],pch=4,col="blue",lwd=3)
points(mu0[1],mu0[2],pch=3,col="red",lwd=3)
legend('topleft',legend=c('95%ellipse','sample means',expression(mu[0])), 
           col=c('orange','blue', 'red'), pch=c(NA,4,3), 
           lwd=c(2,NA, NA), cex=0.7)

```

In the above plot above we can see a visualization of the 95% ellipse confidence, the sample means (blue cross) and the $\mu_0$ (red cross). As it is evident from the plot the $\mu_0$ lie inside the ellipse thus the vector contains the values of the means.

- Because $T^2$ is smaller than the citical value of $T^2$ at $\alpha$ = 5% we can not reject the null hypotheses and we can conclude that the population means of the male birds are plausible means for the female birds.


## b) 

Construct the simultaneous 95% $T^2-intervals$ for $\mu_1$ and $\mu_2$ and the 95% Bonferroni intervals for $\mu_1$ and $\mu_2$. Compare the two sets of intervals. What advantage, if any, do the $T^2-intervals$ have over the Bonferroni intervals?


$T^2~Simultaneous~intervals$
----------------------------

```{r, echo=F}
# Simultaneous Intervals

f <- sqrt(((n-1)*p/(n-p))*qf(1-a, p, n-p))
sim_low <- round((t(xbar) - f * sqrt(diag(S)/n)),2)
sim_up  <- round((t(xbar) + f * sqrt(diag(S)/n)),2)
sim_interval=rbind(sim_low, sim_up)
rownames(sim_interval)=c("lower band", "upper band")
kable(sim_interval, caption = "Simultaneous Intervals Table")

```

Bonferroni intervals
--------------------

```{r, echo=F}
#Bonferroni Intervals
t <- qt((1-a/(2)), df = (n-1))
bon_low <- round((t(xbar) - t * sqrt(diag(S)/n)),2)
bon_up  <- round((t(xbar) + t * sqrt(diag(S)/n)),2)
bon_interval=rbind(bon_low, bon_up)
rownames(bon_interval) <- c("lower band", "upper band")

kable(bon_interval, caption = "Bonferroni Intervals Table")

```



As it is evident, $T^2-simultaneous$ CI is slightly wider than Bonferroni Intervals. Bonferroni method provides shorter intervals
when m = p. Because they are easy to apply and provide the relatively short confidence intervals needed for
inference.

According to the book: "The simultaneous confidence intervals($T^2$) are ideal for “data snooping.” The confidence coefficient $1-\alpha$ remains unchanged for any choice of **a**, so linear combinations of the components $\mu_i$ that merit inspection based upon an examination of the data can be estimated.


## c)
Is the bivariate normal distribution a viable population model? Explain with reference to Q-Q plots and scatter digram

Q-Q plots
---------

```{r, echo=F}
# c) ----------------------------------------------------------------------

# qqnorm(birds[,1], main = "Q-Q plot for x1", 
#        col="purple", pch=19, panel.first=grid(25, 25))
# qqline(birds[,1], col="orange", lwd=2)

qqPlot(birds$`Tail Length`, main ="QQ plot for X1: Tail length",id=F)

```



```{r, echo=F}
# qqnorm(birds[,2], main = "Q-Q plot for x2",
#        col="mediumaquamarine", pch =19,
#        panel.first=grid(25, 25))
# qqline(birds[,2], col="mediumslateblue", lwd=2)

qqPlot(birds$`Wing Length`, main="QQ plot for X2: Wing Length",id=F )

```


Scatter plot
------------

```{r, echo=F}
par(bg='whitesmoke')
plot(birds[,1], birds[,2], 
     xlab=colnames(birds)[1], ylab = colnames(birds)[2],
     col="tomato",pch=19, panel.first = grid(25,25),
     main="Scatter plot wing length~tail length")

```


<p> The Q-Q plot for both variables ($x_1$ & $x_2$) illustrate a linear trend. The scatterplot of two variables also indicates a linear relationship between these two features. These linear trends can lead us to this conclusion that the population can be considered as normal.</p>


# Problem 2

## Data Overview

```{r, echo=F}
# a) ----------------------------------------------------------------------
kable(head(Skulls, n=4))

```



## a)

Plot the data using graphs that you find informative. Justify your choice.

Pair plots
----------

```{r, echo=F, messenge=F, fig.width=12, fig.height=12}

ggpairs(Skulls, mapping = aes(color = epoch)) + theme_bw() 
```


Looking at the different distributions of the variables we can see that all of them are distributed around a mean.
The variables mb and bl seem to be symmetrical while the variables bh and nh seem not. 
Looking at the scatterplots and the corresponding correlations we can observe that apart from bh with mb and nh and bl all of them have a slight
correlation.


## b)

Test (at 5% significance level) if the mean vectors differ for different epochs.


MANOVA
------

```{r, echo=F}
res = manova(cbind(mb,bh,bl,nh)~epoch, Skulls)
print("=================== Result ========================")
res
cat("\n")
cat("\n")
print("=================== Summary =======================")
summary(res)
```

The mean vectors do defer except nasal height. All other means are different between the epochs with a significant level of 5%.
Moreoever, according to the p-value given from *Summary* we should reject the hypothesis that all in-group means (per-epoch means) are equal.

## c)

Construct the 95% simultaneous confidence intervals for the components of the mean vectors.


Confidence intervals
--------------------


```{r, echo=F}

# c) ----------------------------------------------------------------------
w_mb =sum(res$residuals[,1]^2)
w_bh=sum(res$residuals[,2]^2)
w_bl=sum(res$residuals[,3]^2)
w_nh=sum(res$residuals[,4]^2)
w=c(w_mb, w_bh, w_bl, w_nh)
epoch =as.character(unique(Skulls$epoch))
g =length(unique(Skulls$epoch))
p =ncol(Skulls)
n = 150
a = 0.05
C = -qt(a/((p-1)*g*(g-1)), (n-g))* sqrt(2*w/(30*(n-g)))
C_mat =matrix(c(1,1,1,1),4)%*%C


#Calculating the mean values of the samples and the differences between them.
xbar=matrix(0, nrow = 5, ncol = 4, dimnames =list(epoch,names(Skulls[,-1])))
dist=matrix(0, 4,4)

for(i in 2:p){
        for(j in 1:g){
                xbar[j,(i-1)] =mean(Skulls[which(Skulls$epoch==epoch[j]),i])
        }
        for(k in 1:4) {dist[k, i-1] <- xbar[1, i-1]-xbar[k+1, i-1]
        }
}

SI_lower = dist-C_mat
SI_upper = dist+C_mat
e1 =round(t(SI_lower),3)
e2 =round(t(SI_upper),3)
interval =matrix(0, 4,4)
for(i in 1:4){
        for(j in 1:4) {
                interval[i,j] =paste("(",e1[i,j],e2[i,j],")" ,sep = " ")
                }
        }
colnames(interval) =names(Skulls[,-1])
rownames(interval) =c("(epoch1 - epoch2)", "(epoch1 - epoch3)","(epoch1 - epoch4)", "(epoch1 - epoch5)")
knitr::kable(interval)
```


\newpage

# Appendix 

```{r code=readLines(knitr::purl("/home/quartermaine/Courses/Multivariate-Statistical-Methods/labs/Assignment 2/Lab2_report.Rmd",documentation = 1)), echo = T, eval = F}
```




